\section{Proofs for Section~\ref{sec:sec:equivalence-modulo}}

\begin{proof}[Proof of Lemma~\ref{lem:compute-power-series}]

    %J
    
        % The proof is an adaptation of a weighted version of the usual ''DFA to regular expression conversion'' (i.e. weighted automaton to rational power series), with state elimination corresponding to source forgetting.
    
        
        % Define an \emph{internal walk} to be a walk that does not visit any sources,  with the possible exception of the first and last vertices in the walk, which are allowed to be sources. 
        Define the \emph{source number} of a vertex in a $k$-sourced graph to be  $i \in \set{1,\ldots,k}$ if the vertex is the $i$-th source, and $0$ otherwise. For
        $i,j \in \set{0,\ldots,k}$,
        define an  $(i,j)$-walk to be a walk where the first vertex has source number $i$, the last vertex has source number $j$, and the remaining vertices in the walk are not sources. For a $k$-sourced graph $G$, define 
        \begin{align*}
        \varphi_{ij}(G) \in \aalg
        \end{align*}
         to be the power series where the $n$-th term is the number of nonempty $(i,j)$-walks of length $n$. The constant term  of this power series is zero, since we only consider nonempty walks. Define 
        \begin{align*}
        \varphi(G) \in \aalg^{\set{0,\ldots,k}^2}
        \end{align*}
        to be the tuple of all of the values returned by the functions $\varphi_{ij}$. 
        The main observation is that $\varphi$ is an algebra homomorphism, in the following sense.
        \begin{claim}
            For every term operation $f$ 
            in the algebra of $k$-sourced graphs (see the diagram below for its type), there is a term operation $f^\varphi$ in the algebra $\aalg$ (see the diagram for its type) which makes the following diagram commute
            \begin{align*}
            \xymatrix{
                (\text{$k$-sourced graphs})^X 
                \ar[r]^f
                \ar[d]_{\varphi^X}
                &
                 (\text{$k$-sourced graphs})^Y
                \ar[d]^\varphi
                \\
                \aalg^{\set{0,\ldots,k}^2 \times X}
                \ar[r]_{f^\varphi}
                &
                \aalg^{\set{0,\ldots,k}^2 \times Y}
            }
            \end{align*}
        \end{claim}
        Before proving the claim, we use it to finish the proof of the lemma. Suppose that $\Aa$ is a register transducer over the algebra of $k$-sourced graphs. The register transducer  $\Bb$, which is over the algebra $\aalg$, has the same input alphabet, and its registers are triples 
        \begin{align*}
        (i,j,r) \qquad \text{where $i,j \in \set{0,\ldots,k}$ and $r$ is a register of $\Aa$.}
        \end{align*}
         The register updates of $\Bb$ are defined by applying the homomorphism from the claim  to the register updates of $\Aa$. 
        The output register of $\Bb$ is $(0,0,r)$, where  $r$ is the output register of $\Aa$.  From the claim it follows 
        \begin{align*}
        \text{$\Aa$ can output $G$ on input $t$} \qquad \iff \qquad 
        \text{$\Bb$ can output $\varphi_{00}(G)$ on input $t$}.
        \end{align*}
        Since $\varphi_{00}(G)$ is the walk series of $G$ when $G$ has no sources, the lemma follows. It remains to prove the claim.

        \begin{proof} 
            It is enough to prove the claim for the case when $f$ is one of the basic operations in the algebra of $k$-sourced graphs. In particular, $Y$ has one element. Here the proof is similar to conversion of a nondeterministic automaton into a regular expression, as in the Kleene Theorem.
            
            \begin{itemize}
                \item Suppose that $f$ is a constant, which represents a  $k$-sourced graph where all vertices are sources. Then we have 
                \begin{align*}
                \varphi_{ij}(f) = \begin{cases}
                    x & \text{if there is an edge from $i$ to $j$ in $f$}\\
                    0 & \text{otherwise.}
                \end{cases}
                \end{align*}
                \item Suppose that $f$ is the  basic operation which inputs two $k$-sourced graphs and outputs their fusion.  
                An $(i,j)$-walk in the fusion $f(G_1,G_2)$ is either an $(i,j)$-walk in $G_1$ or an $(i,j)$-walk in $G_2$, this is because vertices between endpoints cannot be sources. Furthermore, these cases are disjoint, since we are working with graphs that have parallel edges.
  		This leads to the following equation for fusions:
                \begin{align*}
					\varphi_{ij}(f(G_1, G_2)) = \varphi_{ij}(G_1) + \varphi_{ij}(G_2).
                \end{align*}
                The right hand side uses component-wise addition, which is a basic operation in $\aalg$.
                \item The last case is when $f$ is a unary operation which forgets  source $c \in \set{1,\ldots,k}$.  Let then $G$ be a $k$-sourced graph. A nonempty $(i,j)$-walk in $f(G)$ is either a nonempty $(i,j)$-walk in $G$, or it has the form
                \begin{enumerate}
                    \item a nonempty $(i,c)$-walk in $G$; followed by 
                    \item finitely many (possibly zero) nonempty  $(c,c)$-walks in $G$; followed by
                    \item a nonempty $(c,i)$-walk in $G$.
                \end{enumerate}
                This leads to the following equation:
                \begin{align*}
                \varphi_{ij}(f(G)) = \varphi_{ij}(G)) +  \varphi_{ic}(G) \times (1+ \varphi_{cc}(G)^+)  \times \varphi_{cj}(G).
                \end{align*}                
            \end{itemize}
        \end{proof}
        
\end{proof}



For a rational power series, the embedding is 
\input{pol-aut-equivalence-any-ring}
\subsection{Equivalence for register transducers that compute walk series}\label{sec:decide-power-series}
By Lemmas~\ref{lem:transduction-to-registers} and~\ref{lem:compute-power-series}, the  equivalence problem from Theorem~\ref{thm:path-equivalence} reduces to equivalence for functional \regTsover{\aalg}, where $\aalg$ is the algebra from Lemma~\ref{lem:compute-power-series}. In this section, we complete the proof the theorem by showing that the latter problem is decidable.
\begin{lemma}\label{lem:functionality-decidable-power-series} If $\aalg$ is the algebra from Lemma~\ref{lem:compute-power-series}, then equivalence is decidable for functional \regTsover{\aalg}. 
\end{lemma}
\begin{proof}
	We will reduce to equivalence of \polTsover{\Zrat}, which is shown decidable in Corollary \ref{cor:equivalence-pol-transducers-Zrat}.
	The following equation allows to simulate Kleene plus by division operation; observe that for any $f$ we have $ff^+ + f = f^+$, and in consequence
	$$
	f^+ = \frac{f}{1-f}.
	$$
	This makes \regTover{\aalg} a \polTover{\Zrat} \emph{with division}, i.e. a polynomial transducer with additional feature that register updates can use division operation.
	The rest of the proof is devoted to showing decidability of equivalence of polynomial transducers with division.
	\begin{lemma}
		Equivalence of \polTsover{\Ring} with division is decidable, for a ring with no zero divisors $\Ring$.
	\end{lemma}
	\begin{proof}
		We will adapt the construction of projective line.
		Consider the following representation of elements of $\Ring$: a pair $(x,y)\in \Ring^2$ represents $\frac{x}{y} \in \Ring$. This representation is not unique, and is undefined on some pairs. Then addition, multiplication and, importantly, division of elements of $\Ring$ can be realised as polynomial operations on their representations in $\Ring^2$. This is employed in definition of the following algebra \balg:
		%\begin{align*}
		%	&\frac{x}{y} + \frac{z}{t} = \frac{xt+yz}{yt},\\ 
		%	&\frac{x}{y} \cdot \frac{z}{t} = \frac{xz}{yt},\\
		%	&\left(\frac{x}{y}\right)^+ = \frac{\frac{x}{y}}{1-\frac{x}{y}} = \frac{x}{y-x},
		%\end{align*}
		%and define algebra $\balg$ the following way:
		\algebradefinition{$\Ring \times \Ring$}{$+_B, \cdot_B$ of arity 2, $^+$ of arity 1:
			\begin{align*}
			&(x,y) +_B (z,t) = (xt+yz, yt),\\
			&(x,y)\cdot_B(z,t) = (xz, yt),\\
			&(x,y)^{+_B} = (x, y-x).
			\end{align*}}
		
		Also observe that there is a ''zero-testing'' polynomial function $\eqzeroinalgb$ that satisfies
		$$
		\eqzeroinalgb(x,y) = 0 \iff (x,y) \text { represents } 0,
		$$
		that is, $\eqzeroinalgb(x,y) = x$.
		
		Now, for $\Aa$ being \polTover{\Ring} with division, one can make a \polTover{\balg} $\Bb$, using equations above. Then for each input tree $t$, $\Bb(t)$ represent $\Aa(t)$.
		
		Thus the reduction is as follows: for polynomial transducers with division $\Aa_1, \Aa_2$ take polynomial transducer $\Bb_{1,2}$ corresponding to $\Aa_1-\Aa_2$. Then $\Aa_1 \equiv \Aa_2$
		%\Leftrightarrow
		if and only if 
		$\eqzeroinalgb(\Bb_{1,2}) \equiv 0$,
		that is,
		polynomial transducer $\eqzeroinalgb(\Bb_{1,2})$ obtained by composing output function of $\Bb_{1,2}$ with $\eqzeroinalgb$, is equivalent to $0$.
	\end{proof}	
\end{proof}

